### 2024-06-25 paper/tex/02-overview.tex
edit

Moving on to the examples in \cref{fig:polarization-examples}, we will delve
into several cases that demonstrate the capabilities of type inference within
\fexists.

@@

\paragraph*{Application of a Polymorphic Function}
  A polymorphic function can be applied to a value of a concrete type. In this
  case, the quantified variables are instantiated to make the application
  type-check. For instance, the polymorphic $[[len]]$ can be applied to a list
  of any positive type, and the resulting type will be inferred by the
  algorithm:
  $$[[· ; Φ ⊢ let x = len([one,two,three]); return x : ↑Int]]$$

\paragraph*{Polymorphic Self-application}
  Unlike predicative systems, \fexists permits self-application of polymorphic
  functions. For instance, $[[id]]$ can be applied to itself resulting in a
  suspended polymorphic computation of type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$, which, in
  turn, can be applied to itself:
  $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$

>>>>

Using the examples provided in \cref{fig:polarization-examples}, we now showcase 
the type inference capabilities and restrictions of \fexists

@@

\paragraph*{Application of a Polymorphic Function}
  A polymorphic function can be applied to a value of a concrete type. In this
  case, the explicit type application is not required: the inference algorithm
  automatically instantiates the quantified variables. For instance, the
  polymorphic $[[len]]$ can be applied to a list of any positive type, and the
  resulting type will be inferred by the algorithm:
  $$[[· ; Φ ⊢ let x = len([one,two,three]); return x : ↑Int]]$$

\paragraph*{Polymorphic Self-application.}
  Unlike predicative systems, \fexists permits self-application of polymorphic
  functions. For instance, $[[id]]$ can be applied to itself resulting in a
  suspended polymorphic computation of type $[[↓(∀α⁺.α⁺ → ↑α⁺)]]$, which, in
  turn, can be applied to itself:
  $$[[· ; Φ ⊢ let x = id(id); let y = x(x); return y : ↑↓(∀α⁺.α⁺ → ↑α⁺)]]$$

### 2024-06-25 paper/tex/03-declarative-system.tex
edit

\paragraph*{Variables}
  Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
  the type of a variable from the context. 
  In literature, one can find another version of this rule,
  that enables inferring a type \emph{equivalent}
  to the type from the context. 
  In our case, the inference of equivalent types
  is admissible in general case by \ruleref{\ottdruleDTPEquivLabel}.

\paragraph*{Annotations}
  The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
  and \ruleref{\ottdruleDTPAnnotLabel} use subtyping.
  The annotation is only valid if the
  inferred type is a subtype of the annotation type.

\paragraph*{Abstractions}
  The typing of lambda abstraction is standard. 
  Rule \ruleref{\ottdruleDTtLamLabel} first checks
  that the given type annotating the argument is well-formed,
  and then infers the type of the body in the extended context.
  As a result, it returns an arrow type of function from the
  annotated type of the argument to the type of the body.
  Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
  It extends the type context with the quantifying variable $[[pa]]$ and 
  infers the type of the body. As a result, it returns a polymorphic type
  quantifying the abstracted variable $[[pa]]$ over the type of the body.

\paragraph*{Return and thunk}
  Rules \ruleref{\ottdruleDTReturnLabel} 
  and \ruleref{\ottdruleDTThunkLabel}
  simply add the corresponding shift constructors to the type of the body.

\paragraph*{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. Alternative approaches that do not require normalization
  will be discussed in \cref{sec:explicit-type-application}.
  After the bind, the rule infers the type of the body 
  and checks that it does not use the bound variables so that they do not 
  escape the scope.

\paragraph*{Let binders}
  Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
  binders: we infer the type of the bound value and continue the typing of the
  computation in the extended context.

@@

\paragraph*{Applicative let binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.

@@

\begin{figure}[ht]
  \begin{minipage}[t]{0.65\linewidth}
    \centering
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolP{}
      \end{minipage}
      \begin{minipage}[t]{0.32\linewidth}
      \ottfundefnunpolN{}
      \end{minipage}
    \caption{Type Depolarization \protect\footnotemark}
    \label{fig:depolarization}
  \end{minipage}
  \hspace{0.5cm}
  \begin{minipage}[t]{0.3\linewidth}
    \centering
    \ottfundefnpolarP{}
    \caption{Type Polarization}
    \label{fig:polarization}
  \end{minipage}

\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{rll}
\hline
\multicolumn{2}{c}{\textbf{Elaboration Judgment}} & \textbf{The Soundness Property} \\
\hline
Negative Subtyping & $[[Γ ⊢ iN ≤ iM ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iN| → |iM| ]]$ \\
Positive Subtyping & $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$ \\
Negative Typing & $[[Γ; Φ ⊢ c : iN ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iN|]]$ \\
Positive Typing & $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$ \\
Application Typing & $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ & $[[|Γ| ; |Φ|, x:|iN| ⊢ fe (x ● ft *) : |iM|]]$ \\
\hline
\systemf Typing & $[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$ & $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$ \\
\hline
\end{tabular}
\caption{Soundness of the Elaboration:
  the elaboration judgment on the left implies the typing judgment on the right}
  \label{tab:elaboration-soundness}
\end{table}

>>>>

\subparagraph*{Variables}
    Rule \ruleref{\ottdruleDTVarLabel} allows us to infer
    the type of a variable from the context. 
    In literature, one can find another version of this rule,
    that enables inferring a type \emph{equivalent}
    to the type from the context. 
    In our case, the inference of equivalent types is admissible by \ruleref{\ottdruleDTPEquivLabel}.

\subparagraph*{Annotations}
  The annotation rules \ruleref{\ottdruleDTNAnnotLabel}
  and \ruleref{\ottdruleDTPAnnotLabel} use subtyping.
  The annotation is only valid if the
  inferred type is a subtype of the annotation type.

\subparagraph*{Abstractions}
  The typing of lambda abstraction is standard. 
  Rule \ruleref{\ottdruleDTtLamLabel} first checks
  that the given type annotating the argument is well-formed,
  and then infers the type of the body in the extended context.
  As a result, it returns an arrow type of function from the
  annotated type of the argument to the type of the body.
  Rule \ruleref{\ottdruleDTTLamLabel} infers polymorphic $[[∀]]$-type. 
  It extends the type context with the quantifying variable $[[pa]]$ and 
  infers the type of the body. As a result, it returns a polymorphic type
  quantifying the abstracted variable $[[pa]]$ over the type of the body.

\subparagraph*{Return and thunk}
  Rules \ruleref{\ottdruleDTReturnLabel} 
  and \ruleref{\ottdruleDTThunkLabel}
  simply add the corresponding shift constructors to the type of the body.

\subparagraph*{Unpack}
  Rule \ruleref{\ottdruleDTUnpackLabel} types elimination of $[[∃]]$.
  First, it infers the normalized type of the existential package.
  The normalization is required to fix the order of the quantifying variables
  to bind them. Alternative approaches that do not require normalization
  will be discussed in \cref{sec:explicit-type-application}.
  After the bind, the rule infers the type of the body 
  and checks that it does not use the bound variables so that they do not 
  escape the scope.

\subparagraph*{Let binders}
  Rule \ruleref{\ottdruleDTVarLetLabel} is a standard rule for typing let
  binders: we infer the type of the bound value and continue the typing of the
  computation in the extended context.

@@

\subparagraph*{Applicative let binders}
  Rules \ruleref{\ottdruleDTAppLetLabel} and \ruleref{\ottdruleDTAppLetAnnLabel}
  infer the type of the applicative let binders.
  Both of them infer the type of the head $[[v]]$ 
  and invoke the application typing to infer the type of the application 
  before recursing on the body of the let binder.

@@

\begin{figure}[t]
  \begin{minipage}[t]{0.60\linewidth}
    \centering
      \begin{minipage}[t]{0.30\linewidth}
      \ottfundefnunpolP{}
      \end{minipage}
      \hspace{0.5cm}
      \begin{minipage}[t]{0.30\linewidth}
      \ottfundefnunpolN{}
      \end{minipage}
    \caption{Type Depolarization \protect\footnotemark}
    \label{fig:depolarization}
  \end{minipage}
  \hspace{0.3cm}
  \begin{minipage}[t]{0.3\linewidth}
    \centering
    \ottfundefnpolarP{}
    \caption{Type Polarization}
    \label{fig:polarization}
  \end{minipage}
\end{figure}

@@

\begin{table}[h]
\centering
\begin{tabular}{rll}
\hline
\multicolumn{2}{c}{\textbf{Elaboration Judgment}} & \textbf{The Soundness Property} \\
\hline
Negative Subtyping & $[[Γ ⊢ iN ≤ iM ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iN| → |iM| ]]$ \\
Positive Subtyping & $[[Γ ⊢ iP ≥ iQ ⤳ ft]]$ & $[[|Γ| ; · ⊢ ft : |iQ| → |iP| ]]$ \\
Negative Typing & $[[Γ; Φ ⊢ c : iN ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iN|]]$ \\
Positive Typing & $[[Γ; Φ ⊢ v : iP ⤳ ft]]$ & $[[|Γ| ; |Φ| ⊢ ft : |iP|]]$ \\
Application Typing & $[[Γ ; Φ ⊢ iN ● args ⇒> iM ⤳ fe; fargs]]$ & $[[|Γ| ; |Φ|, x:|iN| ⊢ fe (x ● fargs) : |iM|]]$ \\
\hline
\systemf Typing & $[[fΓ ; fΦ ⊢ ft : fT ⤳ c]]$ & $[[+|fΓ|+ ; +|fΦ|+ ⊢ c : ↑+|fT|+]]$ \\
\hline
\end{tabular}
\caption{Soundness of the Elaboration:
  the elaboration judgment on the left implies the typing judgment on the right}
  \label{tab:elaboration-soundness}
\end{table}
